---
title: "Process GWAS Catalog Variants and output rsIDs"
---

################################################################################################

### STEP 1: Putting all of our files into a single data frame
################################################################################################

```{r}
library(tidyverse)
library(data.table)

#Grab input files
setwd("~/Downloads/cancerproject/") #Update this to the directory with all your files
fnames = list.files(pattern = "^gwas-association-downloaded_.*\\.tsv$")

#Generate data frame with all our data and add a column with fileName
gwas = rbindlist(sapply(fnames, read.delim, simplify = FALSE), use.names = TRUE, idcol = "FileName") %>%
  mutate(
    cancer_type = FileName %>%
      str_remove("^.*withChildTraits_") %>%
      str_remove("\\.tsv$")
  )
colnames(gwas)
```

1. Initial number of SNPs
```{r}
dim(gwas)[1] 
```

2. Initial number of studies
```{r}
length(unique(gwas$PUBMEDID))
```

3. Initial number of subtypes
```{r}
unique(gwas$DISEASE.TRAIT)

```

4. Initial races and/or ethnicities in project

Next step for this will be you guys figuring out how to pull ancestry and 
ethnicity from the INITIAL.SAMPLE.SIZE column

```{r}
summarise_studies = gwas %>%
  group_by(PUBMEDID, MAPPED_TRAIT) %>% #group by the study and trait
  count(INITIAL.SAMPLE.SIZE, name = "n_associations") #output sample composition and the number of observations which is the number of associations
```
################################################################################################

### STEP 2: Clean up data and count up the associated variants per chromosome

################################################################################################

```{r}

#Clean up the messy data
gwas_checked <- gwas %>%
  mutate(
    # Remove allele-annotated rsIDs from SNPS column
    SNPS_clean = str_remove_all(SNPS, "rs[0-9]+-[A-Za-z]+"),
    
    # Remove plain rsIDs (not followed by -) from strongest risk allele column
    `STRONGEST.SNP.RISK.ALLELE` =
      str_remove_all(`STRONGEST.SNP.RISK.ALLELE`,
                     "\\brs[0-9]+\\b(?!-)"),
    
    # Clean up extra delimiters that may result
    `STRONGEST.SNP.RISK.ALLELE` =
      str_replace_all(`STRONGEST.SNP.RISK.ALLELE`, ";\\s*;", ";"),
    
    `STRONGEST.SNP.RISK.ALLELE` =
      str_replace_all(`STRONGEST.SNP.RISK.ALLELE`, "^;\\s*|;\\s*$", "")
  ) %>%
  mutate(
    n_rsids = str_count(SNPS_clean, "rs[0-9]+"),
    n_pos   = str_count(CHR_POS, "[0-9]+")
  ) %>%
  filter(n_rsids == n_pos)

###remove SNPs without known rsIDs, multi-allelic, and clean chromosome IDs
gwas_clean <- gwas_checked %>%
  separate_rows(SNPS, SNP_ID_CURRENT, `STRONGEST.SNP.RISK.ALLELE`, CHR_ID, CHR_POS, sep =  "\\s*x\\s*|;") %>%  # split on " x " or ;
  mutate(SNPS = str_trim(SNPS)) %>%
  filter(str_detect(SNPS, "^rs[0-9]+$")) %>% 
  filter(!grepl("-[ACGT]{2,}$", `STRONGEST.SNP.RISK.ALLELE`)) %>%  #remove obvious multi-allelic risk SNPs
  mutate(
  allele = sub("^.*-", "", `STRONGEST.SNP.RISK.ALLELE`),
  SNP_ID_CURRENT = if_else(
    is.na(SNP_ID_CURRENT) | SNP_ID_CURRENT == "",
    SNPS,
    paste0("rs", SNP_ID_CURRENT)
  ), #add rsID for those split rows or rs to current SNP ID
  IDmatch = if_else(SNPS == SNP_ID_CURRENT, 1L, 0L), #indicator to see if IDs match
  CHR_ID_clean = gsub("[^0-9XY]", "", CHR_ID) #only keep digits or X/Y for chromosome
) %>%
  filter(grepl("^[0-9]+$", CHR_ID_clean)) %>% #final filter to keep autosomal SNPs
  mutate(CHR_ID_clean = as.integer(CHR_ID_clean)) %>% #now make it an integer 
  select(cancer_type, PUBMEDID, INITIAL.SAMPLE.SIZE, DISEASE.TRAIT, MAPPED_TRAIT, MAPPED_GENE, CHR_ID_clean, SNPS, SNP_ID_CURRENT, allele) %>%
  distinct() %>% #remove duplicate rows 
  arrange(-desc(CHR_ID_clean)) #sort chromosomes from 1-22

#write out the processed files with all info and rsIDs
write.table(
    gwas_clean,
    file = "processed_gwas-associations.tsv", #update this with a path if you need
    sep = "\t",
    quote = FALSE,
    row.names = FALSE,
    col.names = TRUE
  )

write.table(
    gwas_clean %>% pull(SNP_ID_CURRENT) %>% unique(), #remove duplicates
    file = "processed_gwas-associations_SNPIDCurrent_rsIDs.txt", #update this with a path if you need
    sep = "\t",
    quote = FALSE,
    row.names = FALSE,
    col.names = FALSE
)
```
################################################################################################

### STEP 3: Pull rsIDs and save to a file split by cancer type

################################################################################################
```{r}
# get unique chromosome + cancer_type combinations
chr_cancer_combos <- gwas_clean %>%
  distinct(CHR_ID_clean, cancer_type)

# loop through each combination
for (i in seq_len(nrow(chr_cancer_combos))) {
  chr_val <- chr_cancer_combos$CHR_ID_clean[i]
  cancer_val <- chr_cancer_combos$cancer_type[i]

  # filter the dataframe
  df_out <- gwas_clean %>%
    filter(CHR_ID_clean == chr_val,
           cancer_type == cancer_val) %>%
    select(SNP_ID_CURRENT) %>%
    distinct()   # unique rsIDs

  # sanitize cancer name for file path (replace spaces with _)
  cancer_file <- gsub(" ", "_", cancer_val)

  # write output file
  write.table(
    df_out,
    #file = paste0(cancer_val, "/chr", chr_val, "_", cancer_file, "_rsids.txt"), # this has a directory structure
    file = paste0("chr", chr_val, "_", cancer_file, "_rsids.txt"), # this will output everything to current directory
    sep = "\t",
    quote = FALSE,
    row.names = FALSE,
    col.names = FALSE
  )
}

```
################################################################################################

### STEP 4: Plotting

################################################################################################
```{r}
cancer_colors <- c(
  "Breast cancer"      = "#27254C",  # JJK: dark indigo
  "Colorectal cancer"  = "#564A94",  # JJK: muted purple
  "Kidney cancer"      = "#9B8AB4",  # JJK: light purple
  "Lung cancer"        = "#F7C83B",  # JJK: golden yellow
  "Ovarian cancer"     = "#000000",  # JJK: black
  "Pancreatic cancer"  = "#39BEB8",  # JJK add: teal accent
  "Prostate cancer"    = "#C1440E"   # FMA: burnt orange (FMA‑style)
)

#count associations per chromosome
chr_counts <- gwas_clean %>%
  group_by(cancer_type) %>% 
  distinct(SNP_ID_CURRENT, .keep_all = TRUE) %>% #remove duplicate rsIDs
  count(CHR_ID_clean, name = "n_associations") %>% #num associations per chrom
  mutate(
    CHR_ID_clean = factor(
      CHR_ID_clean,
      levels = c(as.character(1:22))
    ),
    cancer_type = str_replace_all(cancer_type, "_", " "),
    cancer_type = str_replace_all(cancer_type, "([a-z])([A-Z])", "\\1 \\2"),
    cancer_type = str_to_sentence(cancer_type)
  )

#plot it
associations_acrossCancers = ggplot(chr_counts, aes(x = cancer_type, y = n_associations, fill = cancer_type)) +
  geom_col() +
  facet_wrap(~ CHR_ID_clean, scales = "free_y") +
  scale_fill_manual(values = cancer_colors) +
  theme_bw(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 20, face = "bold"),
    axis.title = element_text(size = 18),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
    axis.text.y = element_text(size = 14),
    strip.text = element_text(size = 16, face = "bold"),
    legend.position = "none"
  ) +
  labs(
    x = "Cancer Type",
    y = "Number of Associations",
    title = "GWAS Associations per Chromosome by Cancer Type"
  )

print(associations_acrossCancers)

```